---
title: "TSA Demo Class: Importing and Cleaning Data Sets"
author: "Luana Lima"
date: "12/05/2022"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

## Loading packages and initializing

First R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output pdf file. 

```{r package, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
```

## Importing Inflow Data

Inflow data for reservoirs in Brazil. 

```{r importdata, message=FALSE, warning=FALSE}

#Importing time series data from text file#
raw_inflow_data <- read.table(file="./Data/inflowtimeseries.txt",header=FALSE,skip=0)

#Trim the table to include only columns you need
nhydro <- ncol(raw_inflow_data)-2
nobs <- nrow(raw_inflow_data)-4 #trim zeros

#If your file does not have header like this one you can add column names after 
#creating the data frame
colnames(raw_inflow_data)=c("Month","Year", "HP1", "HP2","HP3","HP4", "HP5",
                            "HP6","HP7", "HP8","HP9","HP10", "HP11","HP12", 
                            "HP13", "HP14","HP15")

#Creating the date object
#using package lubridate function my()
my_date <- my(paste(raw_inflow_data[1:nobs,1],raw_inflow_data[1:nobs,2],sep="-"))  #function my from package lubridate

#add that to inflow_data and store in a new data frame 
inflow_data <- data.frame("Date"=my_date,"Amazon.River"=raw_inflow_data[1:nobs,"HP15"])

write.csv(inflow_data,file="./Data/Processed/AmazonInflow.csv",row.names = FALSE)

```

## Importing Electricity Price Data

We will work with monthly average for electricity retail price in US. The data is from the U.S. Energy Information Administration and can be download [here][https://www.eia.gov/electricity/data/browser/#/topic/7?agg=2,0,1&geo=g&freq=M%2013:41:41%20GMT-0500%20(EST)]. 

```{r}

#Importing time series data from text file#
electricity_price <- read.csv(file="./Data/Average_retail_price_of_electricity_United_States_monthly.csv",header=TRUE,skip=4)

#Inspect data
head(electricity_price)
nvar <- ncol(electricity_price) - 1
nobs <- nrow(electricity_price)

#Preparing the data - create date object and rename columns
electricity_price_processed <-
  electricity_price %>%
  mutate( Date = my(Month) ) %>% 
  rename( All.sectors = all.sectors.cents.per.kilowatthour ) %>% 
  rename( Residential = residential.cents.per.kilowatthour ) %>% 
  rename( Commercial = commercial.cents.per.kilowatthour ) %>% 
  rename( Industrial = industrial.cents.per.kilowatthour ) %>% 
  arrange( Date ) %>% 
  select( Date, Residential)

head(electricity_price_processed)
summary(electricity_price_processed)

#No NAs so we don't need to worry about missing values

write.csv(electricity_price_processed,file="./Data/Processed/ElectricityPrice.csv",row.names = FALSE)

```
## Importing Renewable Generation Data

Consider the data file "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2022 **Monthly** Energy Review. You will work only with the following column: Total Renewable Energy Production.  


```{r}
#Load/install required package here
library(xlsx)

#Importing data set - using xlsx package
energy_data <- read.xlsx(file="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",header=FALSE,startRow = 13,sheetIndex=1) #startRow is equivalent to skip on read.table

#Now let's extract the column names from row 11 only
read_col_names <- read.xlsx(file="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",header=FALSE,startRow = 11,endRow=11,sheetIndex=1)

colnames(energy_data) <- read_col_names
head(energy_data)

energy_data_processed <- energy_data %>% 
  mutate(Date = Month) %>% 
  mutate(Renewable.Energy = `Total Renewable Energy Production`) %>% 
  select(Date, Renewable.Energy)

write.csv(energy_data_processed,file="./Data/Processed/RenewableEnergy.csv",row.names = FALSE)

```
